
\chapter{Introduction}
\label{ch:intro}

If this marks your first exposure to the new and exciting discipline of
\textit{data science}, you occupy an enviable position. Still in front of you
is all the cool stuff, even the first few sparks of magic when you learn how to
plug data into electrical sockets, perform automated prediction, and write the
first gems of code to probe the depths of an interesting data set. I'm a bit
jealous, tbh, but am also excited to explore it all again with you, which is
the next best thing!

This field has changed the world like hardly any other has, and on an
incredibly short time scale, too. Just a couple decades ago, businesses and
organizations were routinely making major decisions based on gut feelings and
anecdotal observations. Doctors eyeballed sets of symptoms and diagnosed
patients largely based on what conditions they themselves had seen before, or
seen recently. Online sellers gave product recommendations that made sense to
\textit{them}, completely missing patterns and trends that would become
apparent if the characteristics and purchasing patterns of past customers were
taken into account.

Part of the reason decision makers made these suboptimal choices was because it
wasn't yet clear how much punch data science would pack. Another reason was
that the technology wasn't there yet: the processing power and storage capacity
to work with extremely large data sets wasn't commonly available, and of course
the data itself hadn't all been gathered yet. No more! All these parts are here
now. And somewhat incredibly, they're all at your disposal for low (or even no)
cost.

\textbf{This is the era of data science.} If you want to understand and make an
impact on your world, I can honestly think of no better field to dive into than
this one, no matter what your sphere of interest.


% data -> info -> know -> wis

% Applications (sports, Netflix, image recognition, Google, NLP, Coronavirus)

% It's dangerous because people will believe you!
