
\chapter{Loops}
\label{ch:loops}

\index{non-linear}
\index{loop}

It's time for our first look at a \textbf{non-linear} program. Up to now, all
of our Python programs have executed step-by-step, start to finish, like a
metronome, with each line of code getting executed exactly once. That's about
to change. In this chapter, we introduce the concept of a \textbf{loop}, which
is a programming construct that directs lines of code to be executed
\textit{repeatedly}, and out of strict sequence.

\section{The two species of loops}

\index{fixed-iteration (loop)}
\index{variable-iteration (loop)}
\index{counter-controlled (loop)}
\index{condition-controlled (loop)}
\index{for loop@\texttt{for} loop}
\index{while loop@\texttt{while} loop}

Although some programming languages try to dress them up further, there are
really only two fundamental kinds of loops in the world:
\textbf{fixed-iteration loops} and \textbf{variable-iteration
loops}.\footnote{Sometimes these are called \textbf{counter-controlled} and
\textbf{condition-controlled} loops, respectively.} The first kind is simpler
to understand and less error-prone; in most languages (Python included) it is
implemented as a ``\textbf{\texttt{for} loop}.'' The tricker, second kind is
available to programmers as a ``\textbf{\texttt{while} loop}.''

Happily for us, it turns out that \texttt{while} loops don't come up much in
Data Science, at least in the beginning. There are some more advanced
techniques that use them (for instance, optimization methods and threshold
detection) but for us it's going to be \texttt{for} loops that dominate the
landscape. So let's figure out how they work.

\section{A word of caution}

\index{query}

But before we embark, a cautionary note. Some of the things that loops can do
-- especially the early examples -- can also be done using the queries of the
last chapter. For instance, we could use a query to find all the strings in a
\texttt{Series} that begin with the letter \texttt{T}, or we could use a loop
to do the same thing.

Here's the rule: \textbf{if you can do it without a loop, that is always
preferred.} There are two reasons for this. First, it's less code to write, and
less error-prone, to use Pandas' built-in features rather than crafting a loop
yourself. That's why they created those features (like queries) after all.

Second, and ultimately even more important, using a Pandas function is
\textit{much faster} to execute than a loop. The reason has to do with how a
loop is eventually broken down into the little instructions a machine can
understand: when Python runs a loop, it plods through the steps methodically,
whereas the Pandas functions are all pre-baked into a water-cooled rocket
engine that can jet out of the gate.

You don't need to know any of those nitty-gritty details. All you have to
remember is: don't ever resort to using a loop unless you can't figure out how
to do what you want without one. (And unfortunately, there are indeed those
times.)

\section{Iterating through an array}

\index{iterate}

Most often, we'll use a \texttt{for} loop to ``loop through,'' or
``\textbf{iterate} through,'' the contents of an aggregate data variable. This
means that instead of executing a snippet of code \textit{once}, we'll execute
it \textit{once per element of the variable}. This ``once per element'' thing
is what makes the code non-linear.

\index{array}
\index{ndarray@\texttt{ndarray} (NumPy)}

Let's start with the first aggregate data type we learned, a NumPy array.
(I've numbered the lines in this example so I can refer to them in the text
below, but the numbers and colons aren't part of Python.)

\begin{Verbatim}[fontsize=\small,samepage=true,frame=single,framesep=3mm]
 1: villains = np.array(['Jafar','Ursula','Maleficent','Gaston'])
 2: print("Here we go!")
 3: for villain in villains:
 4:     print("Oooo, {} is scary!".format(villain))
 5:     print("({} has {} letters.)".format(villain, len(villain)))
 6: print("Whew!")
\end{Verbatim}

\index{loop!header}
\index{loop!body}

Immediately after creating our \texttt{villains} array, and printing an
introductory message, we encounter our first loop. A loop consists of two
parts: the \textbf{loop header} and the \textbf{loop body}. Here are the rules:

\index{indentation}
\index{curlies (curly braces)}
\index{\{\}@\texttt{\{\}} (curlies)}

\begin{compactitem}
\item The loop header consists of the line that begins with ``\texttt{for}''.
\item The loop body consists of \textit{all of the consecutive following lines
that are \textbf{indented} (tabbed-over) one tab.}\footnote{Other programming
languages -- every other one I know besides Python, in fact -- uses some other
way to designate the loop body than indentation. Many (R and Java, for
instance) use curly braces before and after the loop body so that the computer
knows where it begins and ends. I personally like this feature of Python's, but
there are haters, and the bottom line is you just have to get used to it.}
\end{compactitem}

That second rule turns out to be more important than it seems at first. A very
(\textit{very!}) common error among beginners is to ``mis-indent'' their code
such that their loop body includes more, or less, than they mean it to. So
heads up.

Before we continue, stare at that code above and convince yourself of these two
facts:

\begin{compactitem}
\item[\leftpointright] The loop header is line \textbf{3}.
\item[\leftpointright] The loop body is lines \textbf{4 and 5}. (\textit{Not} line 4
only! \textit{Not} lines 4, 5, and 6!)
\end{compactitem}

Now the reason this is important is that a \texttt{for} loop works as follows:

\index{memory picture}

\begin{compactenum}
\itemsep.1em
\item First, create a new variable (on the left-hand side of the memory
picture) named whatever comes immediately after the word ``\texttt{for}''. (In
this example, the name of this new variable will be \texttt{villain}.)
\item Then, for \textit{each} element of the array, in succession:
    \begin{compactenum}
    \itemsep.1em
    \item Set that variable's value to the element of the array.
    \item \textit{Execute the entire loop body.} (In this example, lines 4--5.)
    \end{compactenum}
\end{compactenum}

